A template for the \emph{arxiv} style

### 2022-09-20 paper-rmd/paper-papaja-revision.Rmd
minor updates

abstract: |
 Bayesian theories of cognitive science hold that cognition is fundamentally probabilistic, but people’s explicit probability judgments often violate the laws of probability. Two recent proposals, the "Probability Theory plus Noise" [@costello.watts2014] and "Bayesian Sampler" [@zhu.etal2020] theories of probability judgments, both seek to account for these biases while maintaining that mental credences are fundamentally probabilistic. These theories fit quite differently into the larger project of Bayesian cognitive science, but their many similarities complicate comparisons of their predictive accuracy. In particular, the models diverge most clearly in their distributional predictions, requiring a careful treatment of errors and external response processes. Here, I cast these theories into a Bayesian data analysis framework that supports principled model comparison using information criteria. Comparing the fits of both models on data collected by Zhu and colleagues [-@zhu.etal2020] I find the data are best explained by an account of biases based on "noise" in the sample-reading process.

@@

keywords          : "probability judgments, Bayesian cognitive science, heuristics and biases"
wordcount         : "9195"

@@

This non-Bayesian account of conditional probabiliity judgments separates the PT+N theory quite fundamentally from the Bayesian Sampler and the larger project of Bayesian cognitive science.

@@

One possibility is that there is another source of error that explains these discrepancies, for instance that people just tend to avoid giving "extreme" responses. Another possibility is that the mental sampling process somehow enforces sampling of both positive and negative examples. A mental sampling process must start somewhere, and these initialization point(s) may influence judgments. For example, features of the initializations of mental sampling processes have been proposed to anchoring [@lieder.etal2012] and unpacking effects [@dasgupta.etal2017; also see @sanborn.chater2016]. In this spirit, one possible explanation for the relative absence of extreme responses is that people draw samples following two initializations, initializing sampling with both a positive and negative instance. As the simulations in Figure 8 (bottom) show, this produces a predicted response distribution that is  approximately $Beta(1 + p N_{\text{eff}}, 1+(1-p)N_{\text{eff}})$.

@@

### Mixture modeling: rounding and contaminants

@@

```{r, echo=FALSE, message=F, warning=F}
texp1 <- read_csv("exp1-trial-level-comparison.csv") %>% mutate(Experiment = c("Exp. 1", "", ""))
texp2 <- read_csv("exp1-trial-level-comparison.csv") %>% mutate(Experiment = c("Exp. 2", "", ""))

@@

The Bayesian adjustment process hypothesized by the Bayesian Sampler model makes two clear predictions about the distribution of participants' responses. First, this implies a truncated range of possible responses. Second, assuming that $\beta$ is constrained to be a relatively small value, then under the Bayesian Sampler model, $N$ influences both the degree to which responses are shrunk toward .50 and the variability of those responses trial-to-trial. Thus, participant's inferred $N$ values should be somehow correlated with the noise in their trial-level responses. In contrast, under the noise-based account of the PT+N model, there is no truncation of responses and no predicted correlation between shrinkage and response variability.

@@

Despite the model's quantitative success, some more qualitative questions remain. First, it bears asking: how plausible are the parameter values inferred from the model? Many participants' estimated $d$ and $d'$ parameters were quite high---potentially against the spirit of the original PT+N model. This model bounds $d$ at .50 in principle, but a sample-reading process with such high error-rates may or may not be plausible. In prior work simulations have often assumed values of $d$ around .10 [e.g. @howe.costello2020; @costello.watts2017]. Further research examining what factors might affect the mental sampling and reading processes (e.g. task complexity, distractions, prior experience) might help to shed light on the plausibility of $d$ estimates in different contexts.

High estimates of $d$ parameters might also call into question arguments for the rational utility of such a process. Zhu and colleagues argue that the regularizing effect of bayesian adjustment should be seen as adaptive. They also consider that "noise" might give an algorithmic-level solution to the computation-level goals defined by the Bayesian Sampler [-@zhu.etal2020]. Even high implied $d$ values might be consistent with rational inference in cases where the number of effective mental samples is very low. For instance, a Beta(2,2) prior is only modestly informative, but could produce $d = .40$ if $N=1$. However, if more samples are drawn then high $d$ values would correspond to potentially inflexible and suboptimal priors (i.e. too large $\beta$ values). From the results, it is clear there are some individuals with both relatively high $d$ and $N$ values, which may press somewhat against the rational justification for the desirability of sampling noise.

As noted, the comparisons of these models using trial-level data rests on a number of elaborating assumptions to support fitting of the models. In particular, the models' implementations incorporate additional assumptions and approximations of the autocorrelated nature of mental samples and of ancillary error processes (e.g. rounding). It should be recognized that different assumptions may have produced different results, and other error processes remain possible. Although other indirect analysis approaches might be designed to avoid these concerns [e.g. @sundh.etal2021], ultimately it seems crucial that cognitive models at some point be fit to the actual human behaviors of interest.

Finally, one important direction for future elaborations of sampling-based theories are more rigorous theories of realistic mental sampling processes, including details of their initialization, autocorrelation, and amortization [@gershman.goodman2016]. Here, I proposed an autocorrelated sampling process initialized with a positive and negative example to account for qualitative features of participants' response distributions and to support a simple approximation. Future work might explore the details of these processes and incorporate them into models of probability judgments in more rigorous ways.

@@

The implications of sampling-based models like the Bayesian Sampler and PT+N theory go well-beyond the probability judgment task itself: these models esssentially rescue probabilistic theories of cognition by rejecting the explicit representation of probabilistic credences. Instead, credences are emergent properties of mental models that probabilistically generate samples [@chater.etal2020].

Clear models of these processes offer the potential to support a measurement model for probabilistic beliefs. Indeed, by representing the "true" subjective probabilities as a latent variable, Bayesian data analysis allows those underlying credences to be inferred. Examining the model posteriors here reveals these estimates often come with considerable uncertainty, but at least for some participants they can be estimated with useful levels of precision. Of course, Zhu and colleagues’ [-@zhu.etal2020] experiments were never designed for this purpose. Future research could explore how estimates of people's credences might be made more reliable, and how inferences about these mental probabilities might be integrated with other Bayesian models of reasoning [e.g. @franke.etal2016; @jern.etal2014; @griffiths.tenenbaum2006]. For instance, people's responses in various reasoning tasks are often explicitly related to inferred subjective mental probabilities [e.g. @jern.etal2014], so accounting for biases in those reports may permit more rigorous model testing. One particularly promising direction could be to integrate these models with formal models of belief revision, which might then shed new light on these fundamental cognitive processes [e.g. @cook.lewandowsky2016; @jern.etal2014; @powell.etal2018].

>>>>

abstract: |
 Bayesian theories of cognitive science hold that cognition is fundamentally probabilistic, but people’s explicit probability judgments often violate the laws of probability. Two recent proposals, the "Probability Theory plus Noise" [@costello.watts2014] and "Bayesian Sampler" [@zhu.etal2020] theories of probability judgments, both seek to account for these biases while maintaining that mental credences are fundamentally probabilistic. These models differ in their averaged predictions about people's conditional probability judgments and in their distributional predictions about their overall patterns of judgments. In particular, the Bayesian sampler's Bayesian adjustment process predicts a truncated range of responses as well as a correlation between the average degree of bias and variability trial-to-trial. However, exploring these distributional predictions with participants' raw responses requires a careful treatment of rounding errors and exogenous response processes. Here, I cast these theories into a Bayesian data analysis framework that supports the treatment of these issues along with principled model comparison using information criteria. Comparing the fits of both models on data collected by Zhu and colleagues [-@zhu.etal2020] I find the data are best explained by an account of biases based on "noise" in the sample-reading process.

@@

keywords          : "probability judgments, Bayesian cognitive science, heuristics and biases"
wordcount         : "9202"

@@

This non-Bayesian account of conditional probability judgments separates the PT+N theory quite fundamentally from the Bayesian Sampler and the larger project of Bayesian cognitive science.

@@

One possibility is that there is another source of error that explains these discrepancies, for instance that people just tend to avoid giving "extreme" responses. Another possibility is that the mental sampling process somehow enforces sampling of both positive and negative examples. Any sampling process must start somewhere, and these initialization point(s) may influence judgments. For example, features of the initializations of mental sampling processes have been proposed to anchoring [@lieder.etal2012] and unpacking effects [@dasgupta.etal2017; also see @sanborn.chater2016]. In this spirit, one possible explanation for the relative absence of extreme responses is that people draw samples following two initializations, initializing sampling with both a positive and negative instance. As the simulations in Figure 8 (bottom) show, this produces a predicted response distribution that is  approximately $Beta(1 + p N_{\text{eff}}, 1+(1-p)N_{\text{eff}})$.

@@

### Mixture modeling: Rounding and contaminants

@@

```{r, echo=FALSE, message=F, warning=F}
texp1 <- read_csv("exp1-trial-level-comparison.csv") %>% mutate(Experiment = c("Exp. 1", "", ""))
texp2 <- read_csv("exp2-trial-level-comparison.csv") %>% mutate(Experiment = c("Exp. 2", "", ""))

@@

The Bayesian adjustment process hypothesized by the Bayesian Sampler model makes two clear predictions about the distribution of participants' responses. First, it implies a truncated range of possible responses. Second, assuming that $\beta$ is constrained to be a relatively small value, then under the Bayesian Sampler model, $N$ influences both the degree to which responses are shrunk toward .50 and the variability of those responses trial-to-trial. Thus, participant's inferred $N$ values should be somehow correlated with the noise in their trial-level responses. In contrast, under the noise-based account of the PT+N model, there is no truncation of responses and no predicted correlation between shrinkage and response variability.

@@

Despite the model's quantitative success, some more qualitative questions remain. First, the plausibility of the parameter values inferred from the model bears consideration. Many participants' estimated $d$ and $d'$ parameters were quite high---potentially against the spirit of the original PT+N model. This model bounds $d$ at .50 in principle, but a sample-reading process with such high error-rates may or may not be plausible. In prior work simulations have often assumed values of $d$ around .10 [e.g. @howe.costello2020; @costello.watts2017]. Further research examining what factors might affect the mental sampling and reading processes (e.g. task complexity, distractions, prior experience) might help to shed light on the most plausible range of $d$ values in different contexts.

High estimates of $d$ parameters might also call into question arguments for the rational utility of such a process. Zhu and colleagues argue that the regularizing effect of Bayesian adjustment should be seen as adaptive. They also consider that "noise" might give an algorithmic-level solution to the computation-level goals defined by the Bayesian Sampler [-@zhu.etal2020]. Even high implied $d$ values might be consistent with rational inference in cases where the number of effective mental samples is very low. For instance, a Beta(2,2) prior is only modestly informative, but could produce $d = .40$ if $N=1$. However, if more samples are drawn then high $d$ values would correspond to potentially inflexible and suboptimal priors (i.e. too large $\beta$ values). From the results, it is clear there are some individuals with both relatively high $d$ and $N$ values, which may press somewhat against the rational justification for the desirability of sampling noise.

Finally as noted, the comparisons of these models using trial-level data rests on a number of elaborating assumptions to support fitting of the models. In particular, the models' implementations incorporate additional assumptions and approximations of the autocorrelated nature of mental samples and of ancillary error processes (e.g. rounding). It should be recognized that different assumptions may have produced different results, and other error processes remain possible. Although other indirect analysis approaches might be designed to avoid these concerns [e.g. @sundh.etal2021], ultimately it seems crucial that cognitive models at some point be fit to the actual human behaviors of interest. An important direction for future elaborations of sampling-based theories are more rigorous theories of realistic mental sampling processes, including details of their initialization, autocorrelation, and amortization [@gershman.goodman2016]. Here, I proposed an autocorrelated sampling process initialized with a positive and negative example to account for qualitative features of participants' response distributions and to support a simple approximation. Future work might explore the details of these processes and incorporate them into models of probability judgments in more rigorous ways.

@@

Probability judgments have proven a fruitful testing ground for sampling-based theories of cognition. But, the implications of sampling-based models like the Bayesian Sampler and PT+N theory go well-beyond the probability judgment task itself: these models have the potential to extend the success of Bayesian theories of cognition to develop a probabilistic science of everyday beliefs. Under such an account, beliefs are not explicitly represented, stored, or even computed as probabilities, but rather they are emergent properties of mental models generating probabilistic samples [@sanborn.chater2016; @chater.etal2020].

Nevertheless we might still use the logic of Bayesian models to understand the operation of these beliefs and how they respond to evidence. Indeed, by representing the "true" subjective probabilities as a latent variable in the models used here, Bayesian data analysis allows those underlying credences to be inferred. Future research could explore how estimates of people's credences might be made more reliable, and how inferences about these mental probabilities might be integrated with other Bayesian models of reasoning [e.g. @franke.etal2016; @jern.etal2014; @griffiths.tenenbaum2006]. For instance, people's responses in various reasoning tasks are often explicitly related to inferred subjective mental probabilities, so accounting for biases in those reports may permit more rigorous model testing. One particularly promising direction could be to integrate these models with formal models of belief revision, which might then shed new light on these fundamental cognitive processes [e.g. @edwards1968; @cook.lewandowsky2016; @jern.etal2014; @powell.etal2018; @powell2022].
\pagebreak
